Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-23 article.tex
DE comments

To a joint probability distribution $P_X$ corresponds a minimal dependency
graph. Each node of this graph is a random variable. Two random variables are
only connected if they are dependent. We consider that two neurons are
connected if  and only if,  there exists  in this dependency  graph an edge
that connects  directly the  two random  variables to  which they correspond.
This is equivalent to writing that two neurons $i$ and $j$ are  not  directly
connected  if and only if  $X_i$  and $X_j$  are  conditionally independent
with  respect to all  other variables. We remind  that two random  variables
$X_i$  and  $X_j$  are  said  to  be  conditionally independent to a (possibly
empty)  set of controlling random variables $Z$, denoted $X_i \perp X_j | Z$,
if the joint conditional probability distribution   $P_{X_i,  X_j   |   Z}$
factorizes  into   $P_{X_i|Z} P_{X_j|Z}$.  In particular,  when  $X_i$ and
$X_j$ are  conditionally independent  with respect  to all  the other
variables, denoted  $X_i \perp  X_j  | X^{-i,j}$,  then  the  observation  of
$X_i$  given  the knowledge  of  $X^{-i,j}$ provides  no  information  about
$X_j$,  and vice-versa.

@@

Our approach implicitly makes the additional assumption  that $P_x$ is  a joint
Gaussian distribution since it  outputs as  degree of association between two
neurons $i$ and $j$ an estimate of $\rho_{X_i, X_j | X^{-i,j}}$. Note also that
the partial correlation statistic is symmetric  (i.e., $\rho_{X_i, X_j |
X^{-i,j}} = \rho_{X_j, X_i | X^{-j,i}}$) and therefore does not account for the
direction of the interaction.

We detail now our procedure for estimating $\rho_{X_i, X_j | X^{-i,j}}$ from
the data. The partial correlation statistics is computed through the inverse
$\Sigma^{-1}$ of the covariance matrix $\Sigma$, also known as the
concentration or precision matrix. Computationally, they can be estimated
efficiently in $O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal
components extracted through PCA~\citep{bishop2006pattern} on the processed
data. In our experiments, $M$ is set to $0.8 p$, thereby allowing us to further
filter the the data by removing the contribution of principal components likely
due to noise.

>>>>

% check dependency map => minimal d-map with distribution?
To a joint probability distribution $P_X$ corresponds a minimal dependency
graph. Each node of this graph is a random variable. Two random variables are
only connected if they are dependent. We consider that two neurons are
connected if  and only if,  there exists  in this dependency  graph an edge
that connects  directly the  two random  variables to  which they correspond.
This is equivalent to writing that two neurons $i$ and $j$ are  not  directly
connected  if and only if  $X_i$  and $X_j$  are  conditionally independent
with  respect to all  other variables. We remind  that two random  variables
$X_i$  and  $X_j$  are  said  to  be  conditionally independent to a (possibly
empty)  set of controlling random variables $Z$, denoted $X_i \perp X_j | Z$,
if the joint conditional probability distribution   $P_{X_i,  X_j   |   Z}$
factorizes  into   $P_{X_i|Z} P_{X_j|Z}$.  In particular,  when  $X_i$ and
$X_j$ are  conditionally independent  with respect  to all  the other
variables, denoted  $X_i \perp  X_j  | X^{-i,j}$,  then  the  observation  of
$X_i$  given  the knowledge  of  $X^{-i,j}$ provides  no  information  about
$X_j$,  and vice-versa.

@@

Our approach implicitly makes the additional assumption  that $P_X$ is  a joint
Gaussian distribution since it  outputs as  degree of association between two
neurons $i$ and $j$ an estimate of $\rho_{X_i, X_j | X^{-i,j}}$. Note also that
the partial correlation statistic is symmetric  (i.e., $\rho_{X_i, X_j |
X^{-i,j}} = \rho_{X_j, X_i | X^{-j,i}}$) and therefore does not account for the
direction of the interaction.

We detail now our procedure for estimating $\rho_{X_i, X_j | X^{-i,j}}$ from
the data. \textcolor{red}{DE finds that this sections is too complicated}
The partial correlation statistics is computed through the inverse
$\Sigma^{-1}$ of the covariance matrix $\Sigma$, also known as the
concentration or precision matrix. Computationally, they can be estimated
efficiently in $O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal
components extracted through PCA~\citep{bishop2006pattern} on the processed
data. In our experiments, $M$ is set to $0.8 p$, thereby allowing us to further
filter the the data by removing the contribution of principal components likely
due to noise.
